```{r}
library(scran)
library(SingleCellExperiment)
library(zellkonverter)
library(MatrixGenerics)
library(BiocParallel)
library(symphony)
library(scDblFinder)
library(transformGamPoi)
library(harmony)
library(dplyr)
library(irlba)
library(ggplot2)
set.seed(0)
```


## NORMALISE TS DATA
## Calculate size factors for all cells with scran -> use Fastcluster clusters for this
```{r}

if (file.exists('../data/ts_blood_vasc_norm.h5ad')){
  print ('Acosh normalisation already done! \n Loading normalised data!')
  ts_dat=zellkonverter::readH5AD(file='../data/ts_blood_vasc_norm.h5ad',reader='python')
  
 } else {
   
  ts_dat=zellkonverter::readH5AD(file='../data/ts_blood_vasc.h5ad',reader='python')
  names(ts_dat@assays)[3]='counts'  
  ## Calculate Size Factors
  ts_dat$fastcluster <- fastcluster(ts_dat)
  input_groups=ts_dat@colData[,'fastcluster',drop=T]
  data_mat=((assay(ts_dat,"counts")))
  register(MulticoreParam(workers = 4))
  size_factors = calculateSumFactors(data_mat, clusters=input_groups)
  rm(data_mat)
  
  ## Apply acosh transformation to the raw counts
  assay(ts_dat, withDimnames=FALSE,"acosh")=acosh_transform(assay(ts_dat, "counts"),size_factors=size_factors,overdispersion=TRUE,
                                                         on_disk = TRUE)
  
  names(ts_dat@assays)[3]='raw_counts'
  ## Save file
  writeH5AD(ts_dat,file='../data/ts_blood_vasc_norm.h5ad')}
```

## Prepare data necessary for Harmony batch correction of TS data
## - Extract metadata of TS data
## - Select veriable genes 
## - Calculate PCA of TS data 
```{r}
## Convert metadata to factors 
meta_ts=ts_dat@colData

## Convert metadata to factors 
meta_ts=ts_dat@colData
#rm(ts_dat)

cols_to_factors_ts=c( "method", "donor")
meta_ts_=as.data.frame(meta_ts) %>%
          mutate_at(vars(cols_to_factors_ts), as.factor)
  

   
## Extract normalised values of the reference data
ref_exp_full=as(assay(ts_dat, "acosh"), "dgCMatrix")

var_genes=vargenes_vst(ref_exp_full, groups = as.character(meta_ts_[['donor']]), topn = 10000)
ref_exp=ref_exp_full[var_genes, ]


## Calculate means and Std.Dev of the genes expressions
vargenes_means_sds = tibble(symbol = var_genes, mean = Matrix::rowMeans(ref_exp))
vargenes_means_sds$stddev = rowSDs(ref_exp, vargenes_means_sds$mean)
ref_exp_scaled = scaleDataWithStats(ref_exp, vargenes_means_sds$mean, vargenes_means_sds$stddev, 1)
  
## Calculate PCA of scaled expression values and extract loadings
set.seed(0)
s = irlba(ref_exp_scaled, nv = 30)
Z_pca_ref = diag(s$d) %*% t(s$v) # [pcs by cells]
loadings = s$u


saveRDS((ref_harmObj), '../data/ts_harmonized_pcs.rds')
saveRDS(s, '../data/ref_exp_scaled_PCA.rds')
saveRDS(vargenes_means_sds, '../data/vargenes_means_sds.rds')
```


## Build reference symphony object from the TS harmonized data
## 1. Run harmony Batch correction to get TS data Harmony object
## 2. Create reference Symphony object from TS data Harmony object
```{r}
### Load previously created and saved PCA loading + MEtadata + Gene variance/mean dataframe
s=readRDS('../data/ref_exp_scaled_PCA.rds')
loadings=s$u
vargenes_means_sds=readRDS('../data/vargenes_means_sds.rds')

## Run harmony batch correction
ref_harmObj = harmony::HarmonyMatrix(
        data_mat = t(Z_pca_ref),  ## PCA embedding matrix of cells
        meta_data = meta_ts_, ## dataframe with cell labels
        vars_use = cols_to_factors_ts,    ## variable to integrate out
        return_object = TRUE,     ## return the full Harmony model object
        do_pca = FALSE            ## don't recompute PCs
        )


# Compress a Harmony object into a Symphony reference
reference = buildReferenceFromHarmonyObj(
                        ref_harmObj,  # output object from HarmonyMatrix()
                        meta_ts_,           # reference cell metadata
                        vargenes_means_sds,     # gene names, means, and std devs for scaling
                        loadings,               # genes x PCs matrix
                        verbose=TRUE,         # verbose output
                        do_umap=TRUE,         # set to TRUE to run UMAP
                        save_uwot_path='./testing_uwot_model_1'
                        )

saveRDS(reference, '../data/ts_reference_object.rds')

```

## NORMALISE QUERY (OUR) DATA
## Calculate size factors for all cells with scran -> use Fastcluster clusters for this
```{r}
dat=readH5AD('../data/data_QC_all_genes.h5ad')

if ('acosh' %in% names(dat@assays)){
  print ('Acosh normalisation already done!')
  } else {
  
  names(dat@assays)='counts'
  dat$fastcluster <- fastcluster(dat)
  input_groups=dat@colData[,'fastcluster',drop=T]
  data_mat=((assay(dat)))
  register(MulticoreParam(workers = 4))
  size_factors = calculateSumFactors(data_mat, clusters=input_groups)
  rm(data_mat)
  
  ## Apply acosh transformation to the raw counts
  assay(dat, withDimnames=FALSE,"acosh")=acosh_transform(assay(dat, "counts"),size_factors=size_factors,overdispersion=TRUE,
                                                          on_disk = TRUE)
  
  names(dat@assays)[1]='X'
  
  ## Save file
  writeH5AD(dat,file='../data/data_QC_all_genes.h5ad')}
```
##  MAP OUR DATA (QUERY) TO REFERENCE (TS DATA)
```{r}

## Convert metadata to factors
meta=dat@colData

cols_to_factors=c( "patient", "batch")
meta_=as.data.frame(meta) %>%
          mutate_at(vars(cols_to_factors), as.factor)

# Map query to reference dataset
query=mapQuery(assay(dat,'acosh'),    # query gene expression (genes x cells)
                 meta_,                 # query metadata (cells x attributes)
                 reference,             # Symphony reference object
                 vars=cols_to_factors,  # Query batch variable(s) to integrate over (column names in metadata)
                 do_normalize = FALSE,  # perform log(CP10k) normalization on query
                 do_umap = TRUE)        # project query cells into reference UMAP

## Run KNN prediction of cell types
query = knnPredict(query,       # query object
                   reference,   # reference object
                   reference$meta_data$free_annotation, # reference cell labels for training
                   k = 5,       # number of reference neighbors to use for prediction
                   confidence = TRUE)

write.csv(query$meta_data,'../data/symphony_transferred_labels.csv')
```


## LOAD SYMPHONY OBJECTS COMPUTED ON THE CLUSTER
```{r}
reference=readRDS('../data/ts_reference_object.rds')
query=readRDS('../data/symph_query_object.rds')
```

## Check knn prediction probablity distribution to set a threshold for certain/uncertainly transferred labels
```{r}
hist(query$meta_data$cell_type_pred_knn_prob,breaks=20)
unc_thr=0.8

## Init 2 new columns -> certain and uncertain
query$meta_data$certain=NA
query$meta_data$uncertain=NA

## For cells where knn_probability is >=unc_thr -> add cell label to 'certain' column
query$meta_data[query$meta_data$cell_type_pred_knn_prob>=unc_thr,'certain']=sapply(query$meta_data[,'cell_type_pred_knn'][query$meta_data$cell_type_pred_knn_prob>=unc_thr],as.character)

## For cells where knn_probability is < unc_thr -> add cell label to 'uncertain' column
query$meta_data[query$meta_data$cell_type_pred_knn_prob<unc_thr,'uncertain']=sapply(query$meta_data[,'cell_type_pred_knn'][query$meta_data$cell_type_pred_knn_prob<unc_thr],as.character)
```


## Plot transferred labels
```{r echo=FALSE,fig.width=7,fig.height=4}
# Sync the column names for both data frames
reference$meta_data$cell_type_pred_knn=reference$meta_data$free_annotation
reference$meta_data$cell_type_pred_knn_prob=NA
reference$meta_data$ref_query = 'reference'
query$meta_data$ref_query = 'query'

coln=c('cell_type_pred_knn','cell_type_pred_knn_prob','ref_query')

# Add the UMAP coordinates to the metadata
meta_data_combined = rbind(query$meta_data[,coln], reference$meta_data[,coln])
umap_combined = rbind(query$umap, reference$umap$embedding)
umap_combined_labels = cbind(meta_data_combined, umap_combined)

# Plot UMAP visualization of all cells
ggplot(data=umap_combined_labels,aes(x=UMAP1,y=UMAP2,colour=cell_type_pred_knn_prob)) + 
       geom_point(size=0.2) +facet_grid(. ~ ref_query)
ggplot(data=umap_combined_labels,aes(x=UMAP1,y=UMAP2,colour=cell_type_pred_knn)) +
       geom_point(size=0.8)+facet_grid(. ~ ref_query)

query_umap=cbind(query$umap,query$meta_data)
ggplot(data=query_umap,aes(x=UMAP1,y=UMAP2,colour=cell_type_pred_knn)) + 
       geom_point(size=0.8) 
ggplot(data=query_umap,aes(x=UMAP1,y=UMAP2,colour=patient)) + 
       geom_point(size=0.8) 
ggplot(data=query_umap,aes(x=UMAP1,y=UMAP2,colour=cell_type_pred_knn_prob)) + 
       geom_point(size=0.8) 

```