

```{r}
library(DESeq2)
library(muscat)
library(SingleCellExperiment)
library(zellkonverter)
```

```{r}
dat=readH5AD('../data/data_scvi_batch_corrected.h5ad')
meta=read.csv('../data/adata_obs_with_metadata.csv')
```

# Determine marker genes of Leiden-clusters of scVI-batch corrected data

## 1. For each Leiden-cluster do:
## 2. Calculate Pseudo-bulk counts (sum counts per each patient)
## 3. Split __sce__ object into 2 datasets: cells in Leiden-clust vs. rest
## 4. Merge the created datasets' counts and create a "coldata" dataframe for creating a design matrix later
## 5. Perform DGE with DESeq2 between Leiden-cluster cells vs. rest, controlling for "patient" as batch effect

```{r}

## Create a coldata dataframe containing the batch effect descriptors
coldata=aggregate(x=meta[,c('preparation_kit','batch','Age')],
                             by=list(patient=meta$patient),FUN=unique)
rownames(coldata)=coldata$patient


## Function for searching cluster marker genes
get_leiden_marker_genes=function(dat,leiden_clust_colname,coldata){
  
  ## Init list for collecting results
  de_result_list=list()
  
  ## Loop over the Leiden-clusters
  for (leiden_clust in unique(sort(dat@colData[,leiden_clust_colname]))){
    print(leiden_clust)
    
    
    ### Split SCE object into cells of Leiden-cluster vs. rest
    leid_clust_dat=dat[,dat@colData[,leiden_clust_colname]==leiden_clust]
    rest_dat=dat[,dat@colData[,leiden_clust_colname]!=leiden_clust]
    
    
    ## Sum the cells per each patient, creating a pseudo-bulk dataset of num_genes x num_patients
    leid_clust_pb=aggregateData(leid_clust_dat, assay="counts", fun='sum', by=c('patient'))
    rest_pb=aggregateData(rest_dat, assay="counts", fun='sum', by=c('patient'))
    
    ## Add suffix to colnames, indicating where the data comes from (cluster or rest dataset)
    leid_clust_pb_coln=paste(colnames(assay(leid_clust_pb)), "_clust", sep = "")
    rest_pb_coln=paste(colnames(assay(rest_pb)), "_rest", sep = "")
    
    ## Reorder coldata to align with count data
    coldata_clust=coldata[colnames(assay(leid_clust_pb)),]
    coldata_rest=coldata[colnames(assay(rest_pb)),]
    
    
    
    ### Concatenate the two count datasets (cluster and rest) + update its colname
    contr_data=cbind(assay(leid_clust_pb),assay(rest_pb))
    contr_data_coln=c(leid_clust_pb_coln,rest_pb_coln)
    colnames(contr_data)=contr_data_coln
    
    
    
    ### Concatenate the coldata as well fpr the clust and rest samples
    contr_data_coldata=rbind(coldata_clust,coldata_rest)
    rownames(contr_data_coldata)=contr_data_coln
    
    ## Add column indicating origin of data (clust or rest) + set rest as level 0
    contr_data_coldata[,'contrast']=c(rep('clust',length(leid_clust_pb_coln)),
                                          rep('rest',length(rest_pb_coln)))
    contr_data_coldata$contrast=factor(contr_data_coldata$contrast, levels=c('rest','clust'))
    
    ## Convert character columns to factors
    cols_to_factors=c('preparation_kit','batch','patient')
    contr_data_coldata[, cols_to_factors] <- lapply(contr_data_coldata[, cols_to_factors], factor)
    
    ## Reorder coldata's rows to match the merged count matrix's columns (pre-requisite for DESeq2)
    contr_data_coldata=contr_data_coldata[colnames(contr_data),]
    
    
    
    ### Perform DGE and extract results
    dge=DESeqDataSetFromMatrix(contr_data, colData=contr_data_coldata, 
                                #design=~1+batch+patient+preparation_kit+contrast)
                               design=~1+patient+contrast)
    dds=DESeq(dge)
    res=results(dds, contrast=c("contrast","clust","rest"))
    deseq_out=as.data.frame(res@listData, row.names = res@rownames)
    
    
    df=deseq_out[,c("log2FoldChange","pvalue","padj")]
    df$gene=rownames(deseq_out)
    colnames(df)=c("log2FC","pvalue","padj","gene")
    df[,'cluster']=leiden_clust
    
    ## Extract only significant genes (FDR<0.05) + sort by most significant
    df=df[(df$padj<0.05)&(!is.na(df$padj)),]
    df=df[order(df$padj),]
    
    ## Add result to result list
    de_result_list[[leiden_clust]]=df
  }
  
  return (de_result_list)  
}
```


## Perform the marker gene search for all genes
```{r}
## Run marker gene function
#colnames(dat@colData)
leiden_clust_colname='leiden_scVI_batch_patient_corr_model'
marker_df_list=get_leiden_marker_genes(dat,leiden_clust_colname,coldata)

## Use do.call to apply rbind to the list of data frames
concatenated_df <- do.call(rbind, marker_df_list)
## Reset rownames
rownames(concatenated_df) <- NULL

## Save concatenated data frame
deseq2_dirname='../data/dge_deseq2'
fname='scvi_leiden_clust_marker_genes.csv'
file_path=file.path(deseq2_dirname,fname)
write.csv(concatenated_df,file=file_path)

```


## Perform the markers genes search for data that is subsetted to HVG (HVG previously calculated in scanpy)
```{r}

## Leiden resolutions (the higher number, the more clusters there are)
leiden_resolutions=c(1)

## List of HVG numbers to loop over
hvg_num_list=c(500,800,1000,2000,4000)

## Dirname for saving the results to
deseq2_dirname='../data/dge_deseq2'


for(leiden_resolution in leiden_resolutions){
  
  for (n_top_genes in hvg_num_list[5:5]){
    
    ## Extract column name containing HVG genes
    hvg_genes_coln=(paste0('highly_variable_',n_top_genes))
    
    ## Extract column name containing Leiden-clusters
    leiden_clust_colname=paste(c('leiden_hvg',n_top_genes,'scVI',leiden_resolution),collapse='_')
    
    ## Subset data to the HVG genes
    dat_hvg=dat[dat@rowRanges@elementMetadata[,hvg_genes_coln],]
    
    ## Perform marker genes search for subsetted HVG data
    marker_df_list=get_leiden_marker_genes(dat_hvg,leiden_clust_colname,coldata)
    
    ## Use do.call to apply rbind to the list of data frames
    concatenated_df <- do.call(rbind, marker_df_list)
    ## Reset rownames
    rownames(concatenated_df) <- NULL
    
    
    ## Save results
    fn=paste(c('scvi_leiden_clust_marker_genes_hvg',n_top_genes,leiden_resolution),collapse='_')
    file_path=file.path(deseq2_dirname, paste0(fn,'.csv'))
    write.csv(concatenated_df,file=file_path)
    
  }
}  
```

## Save list of DGE result dataframes

```{r}
unique(dat@colData[,'leiden_scVI'])

unique(concatenated_df$cluster)
```

